---
title: "DTS 301 — Machine Learning I"
subtitle: "Week 0: R for Data Science, Data Manipulation, Visualization & Exploratory Data Analysis"
author: |
  **Dr. Sakinat Folorunso**  
  Associate Professor of AI Systems  
  Department of Computer Science, Olabisi Onabanjo University (OOU)  
  Email: sakinat.folorunso@oouagoiwoye.edu.ng
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Week 0 — Student Guide

## Purpose

This practical week provides the **R-language pathway** for the foundation of DTS 301.

Before we ask a computer to **learn from data**, we need to be comfortable with:

- R programming;
- vectors and matrices;
- data frames;
- data manipulation;
- visualization;
- exploratory data analysis (EDA).

The learning cycle is:

> **Read → Predict → Run → Inspect → Explain → Reflect**

Do not rush through the code. Read the explanation, predict what will happen, run the code, inspect the output, and explain what you observed.

### Course progression

**Week 0 → R for Data Science & EDA**  
**Week 1 → Introduction to Machine Learning**  
**Week 2 → Data Representation, Linear Transformations & Matrix–Vector Operations**

> **Important:** The goal is not simply to make the code run. The goal is to understand what the code is doing and why it matters for machine learning.

---

# 1. Getting Started with RStudio

## What is RStudio?

RStudio is an integrated development environment (IDE) that makes it easier to write, run, organise and document R code.

A typical RStudio window contains four main areas:

1. **Source pane** — where you write scripts or R Markdown documents.
2. **Console** — where R commands are executed.
3. **Environment/History** — where you can inspect objects created during a session.
4. **Files/Plots/Packages/Help/Viewer** — where you manage files, view plots, inspect packages and access documentation.

## Your first RStudio task

Create a new R Markdown document:

**File → New File → R Markdown**

Save it as:

`DTS301_Week00_R_for_Data_Science_Student_Centered.Rmd`

You will work through this document by running individual code chunks.

### Useful shortcuts

- **Run current line/selection:** `Ctrl + Enter`
- **Run current code chunk:** `Ctrl + Shift + Enter`
- **Knit the document:** use the **Knit** button at the top of the Source pane.

> **Good practice:** Save your work regularly.

---

# 2. R Objects and Data Types

R stores information in objects.

We begin with:

- numeric values;
- character values;
- logical values;
- vectors;
- basic arithmetic.

The aim is to build enough R fluency to work confidently with data.

## Worked Example

```{r}
# Store the number 25 in an object called age.
age <- 25

# Display the value stored in age.
print(age)

# Display the data type of age.
print(class(age))
```

## Predict

Before running the code:

1. What will the first output be?
2. What type of object is `age`?

### Explain

In your own words, what is an **object** in R?

---

## More Data Types

```{r}
# Store a student's name as a character value.
student_name <- "Amina"

# Store the student's examination score as a numeric value.
score <- 78

# Store the student's attendance as a decimal value.
attendance <- 0.85

# Store whether the student passed as a logical value.
passed <- TRUE

# Display the student's name.
print(student_name)

# Display the examination score.
print(score)

# Display the attendance value.
print(attendance)

# Display the logical value.
print(passed)

# Display the class of the student's name.
print(class(student_name))

# Display the class of the score.
print(class(score))

# Display the class of the attendance value.
print(class(attendance))

# Display the class of the logical value.
print(class(passed))
```

---

# 3. Basic Arithmetic in R

R can perform common arithmetic operations.

| Operator | Meaning |
|:--:|---|
| `+` | Addition |
| `-` | Subtraction |
| `*` | Multiplication |
| `/` | Division |
| `^` | Exponentiation |
| `%%` | Remainder |

These operations form the foundation for numerical data analysis.

## Worked Example

```{r}
# Store the first test score.
test_1 <- 72

# Store the second test score.
test_2 <- 84

# Add the two scores.
total <- test_1 + test_2

# Calculate the average score.
average <- total / 2

# Display the total score.
print(total)

# Display the average score.
print(average)
```

## Do It Yourself

Create three variables representing three assessment scores.

Calculate:

- the total;
- the mean;
- the difference between the highest and lowest score.

**Challenge:** Change the values and observe how the results change.

```{r}
# Store your first assessment score.
score_1 <- 70

# Store your second assessment score.
score_2 <- 80

# Store your third assessment score.
score_3 <- 90

# Calculate the total of the three scores.
total_score <- score_1 + score_2 + score_3

# Calculate the mean of the three scores.
mean_score <- total_score / 3

# Find the difference between the largest and smallest score.
score_range <- max(score_1, score_2, score_3) - min(score_1, score_2, score_3)

# Display the total.
print(total_score)

# Display the mean.
print(mean_score)

# Display the range.
print(score_range)
```

---

# 4. Vectors in R

A vector stores multiple values of the same basic type.

Vectors are fundamental in R because they allow us to work with collections of observations efficiently.

For example, a vector can contain:

- examination scores;
- ages;
- temperatures;
- study hours.

R uses **one-based indexing**, meaning the first element is at position `1`.

## Worked Example

```{r}
# Create a vector containing five examination scores.
scores <- c(65, 72, 81, 76, 90)

# Display the complete vector.
print(scores)

# Display the first element using position 1.
print(scores[1])

# Display the third element using position 3.
print(scores[3])

# Display the first three elements.
print(scores[1:3])

# Add a new score to the vector.
scores <- c(scores, 88)

# Display the updated vector.
print(scores)
```

## Think

Why is the first element accessed using `scores[1]` in R?

Compare this with Python, where indexing starts at `0`.

## Do It Yourself

Create a vector containing five values from a dataset of your choice.

Then:

1. display the first value;
2. display the last value;
3. display the first three values;
4. add one new value;
5. display the updated vector.

```{r}
# Create a vector containing five values of your choice.
my_values <- c(10, 20, 30, 40, 50)

# Display the first value.
print(my_values[1])

# Display the last value.
print(my_values[length(my_values)])

# Display the first three values.
print(my_values[1:3])

# Add a new value to the vector.
my_values <- c(my_values, 60)

# Display the updated vector.
print(my_values)
```

---

# 5. Matrices in R

A matrix is a two-dimensional numerical structure arranged into rows and columns.

Matrices are especially important in machine learning because datasets are often represented as:

- rows = observations;
- columns = features.

This idea will become central in Week 2.

## Worked Example

```{r}
# Create a matrix with three observations and four features.
X <- matrix(
  c(80, 75, 90, 85,
    60, 65, 70, 68,
    92, 88, 95, 90),
  nrow = 3,
  byrow = TRUE
)

# Display the matrix.
print(X)

# Display the dimensions of the matrix.
print(dim(X))

# Display the number of rows.
print(nrow(X))

# Display the number of columns.
print(ncol(X))
```

## Understanding Dimensions

If a matrix has dimensions:

```text
3 × 4
```

that means:

- 3 rows;
- 4 columns.

In a machine-learning dataset, this could represent:

> **3 observations described by 4 features.**

## Indexing a Matrix

```{r}
# Display the first row of X.
print(X[1, ])

# Display the second row of X.
print(X[2, ])

# Display the first column of X.
print(X[, 1])

# Display the value in the first row and first column.
print(X[1, 1])

# Display the first two rows and first two columns.
print(X[1:2, 1:2])
```

---

# 6. Lists and Data Frames

A **list** can contain different types of objects.

A **data frame** is a tabular structure in which columns can contain different types of data.

For data science, data frames are particularly important because they closely resemble ordinary datasets or spreadsheets.

## Lists

```{r}
# Create a list containing different types of information.
student <- list(
  name = "Amina",
  attendance = 85,
  test_score = 78,
  passed = TRUE
)

# Display the complete list.
print(student)

# Retrieve the student's name using its named element.
print(student$name)

# Retrieve the student's test score.
print(student$test_score)
```

---

# 7. Creating a Data Frame

We can represent several student observations as a data frame.

Think of:

- each row as an observation;
- each column as a variable;
- some variables as possible features;
- one variable as a possible target.

This is the bridge from programming to data analysis.

```{r}
# Create a data frame containing student information.
df <- data.frame(
  student = c("Amina", "Bola", "Chidi", "David", "Efe"),
  attendance = c(85, 70, 92, 60, 78),
  test_score = c(78, 65, 88, 55, 72),
  final_score = c(82, 68, 91, 58, 75)
)

# Display the complete data frame.
print(df)
```

---

# 8. Inspecting a Dataset

Before modelling, we need to understand our data.

Useful R functions include:

- `head()` — first rows;
- `tail()` — last rows;
- `dim()` — rows and columns;
- `str()` — structure and data types;
- `summary()` — summary statistics;
- `names()` — column names.

```{r}
# Display the first five rows of the data frame.
print(head(df))

# Display the dimensions of the data frame.
print(dim(df))

# Display the column names.
print(names(df))

# Display the internal structure of the data frame.
str(df)

# Display summary statistics.
print(summary(df))
```

## Inspect and Explain

Look at the output and answer:

1. How many observations are present?
2. How many columns are present?
3. Which columns are numerical?
4. Which column identifies the student?
5. What is the average final score?
6. What is the range of attendance values?

The goal is to **interpret** the dataset, not simply execute the commands.

---

# 9. Selecting Columns and Rows

R allows us to select particular variables and observations.

A column can be selected using:

```r
df$attendance
```

or:

```r
df[["attendance"]]
```

Rows can be filtered using logical conditions.

This becomes important when exploring subsets of a dataset.

```{r}
# Select the attendance column.
attendance <- df$attendance

# Display the attendance values.
print(attendance)

# Select the attendance and final score columns.
selected_columns <- df[, c("attendance", "final_score")]

# Display the selected columns.
print(selected_columns)

# Select students whose attendance is at least 80.
high_attendance <- df[df$attendance >= 80, ]

# Display the filtered observations.
print(high_attendance)
```

---

# 10. Feature Engineering

Feature engineering means creating or transforming variables so that information becomes more useful for analysis or modelling.

For example, we can create a combined score from existing variables.

The key idea is:

> **A feature is a representation of information that a model may use as input.**

Feature engineering should be guided by the problem and by what information would genuinely be available at prediction time.

```{r}
# Create a combined score using attendance and test score.
df$combined_score <- (df$attendance + df$test_score) / 2

# Display the updated data frame.
print(df)
```

## Do It Yourself — Feature Engineering

Create a new variable called `performance_index`.

Use:

- 40% of `test_score`;
- 60% of `final_score`.

Then inspect the new variable.

**Think:** Why might a researcher create a new feature instead of using the original variables directly?

```{r}
# Create the performance index using 40% of the test score.
df$performance_index <- 0.4 * df$test_score + 0.6 * df$final_score

# Display the updated data frame.
print(df)
```

---

# 11. Missing Data

Real datasets are rarely perfect.

Common problems include:

- missing values;
- inconsistent entries;
- duplicated observations;
- incorrect data types;
- unusual values.

A missing value in R is usually represented as `NA`.

Before modelling, we need to detect and investigate missing values.

```{r}
# Create a data frame containing one missing test score.
missing_data <- data.frame(
  student = c("Amina", "Bola", "Chidi"),
  test_score = c(78, NA, 88),
  final_score = c(82, 68, 91)
)

# Display the data frame containing the missing value.
print(missing_data)

# Count missing values in each column.
print(colSums(is.na(missing_data)))
```

## Handling Missing Values

There is no universally correct way to handle missing data.

Possible approaches include:

- removing observations;
- removing variables;
- imputing missing values;
- investigating why the values are missing.

The correct approach depends on the context and modelling task.

For now, the important skill is to **detect missingness before modelling**.

```{r}
# Calculate the mean of the available test scores while ignoring the missing value.
mean_test_score <- mean(missing_data$test_score, na.rm = TRUE)

# Replace the missing test score with the calculated mean.
missing_data$test_score[is.na(missing_data$test_score)] <- mean_test_score

# Display the completed data frame.
print(missing_data)
```

### Important Question

Is replacing a missing value with the mean **always** appropriate?

No.

Before choosing a method, ask:

- Why is the value missing?
- How much data is missing?
- Could missingness itself contain information?
- Could the method introduce bias?
- Is the information available legitimately?

This is part of responsible data analysis.

---

# 12. Visualization with Base R

Visualizations can reveal patterns that are difficult to see in tables.

We will use base R to create:

- histograms;
- scatter plots;
- boxplots.

Visualization is part of EDA, not decoration.

## Histogram

```{r}
# Create a vector of examination scores.
scores <- c(55, 60, 65, 70, 72, 75, 78, 80, 82, 85, 88, 90, 92, 95)

# Create a histogram of the examination scores.
hist(scores,
     main = "Distribution of Examination Scores",
     xlab = "Examination Score",
     ylab = "Number of Students")
```

## Interpreting a Histogram

Ask:

- Where are most observations concentrated?
- Is the distribution approximately symmetric?
- Are there possible extreme values?
- Is the variable spread out or tightly clustered?

The data-science skill is not merely creating a plot.

> **The skill is explaining what the plot tells you about the data.**

## Scatter Plot

```{r}
# Create a vector representing study hours.
study_hours <- c(2, 3, 4, 5, 6, 7, 8, 9, 10, 11)

# Create a vector representing examination scores.
exam_scores <- c(50, 55, 60, 65, 68, 72, 78, 82, 88, 91)

# Create a scatter plot of study hours against examination scores.
plot(study_hours,
     exam_scores,
     main = "Study Hours vs Examination Score",
     xlab = "Study Hours",
     ylab = "Examination Score")
```

### Think

What relationship appears to exist between study hours and examination score?

Be careful:

> A visible relationship does not automatically prove that one variable causes the other.

This distinction will matter throughout data science and machine learning.

## Boxplot

```{r}
# Create a boxplot of the examination scores.
boxplot(exam_scores,
        main = "Distribution of Examination Scores",
        ylab = "Examination Score")
```

---

# 13. Correlation

Correlation measures the strength and direction of a **linear relationship** between numerical variables.

A correlation coefficient is commonly represented by:

$$r \in [-1,1]$$

Very roughly:

- values near `+1` indicate a strong positive linear relationship;
- values near `-1` indicate a strong negative linear relationship;
- values near `0` indicate little linear relationship.

Correlation does **not** establish causation.

```{r}
# Calculate the correlation coefficient between study hours and examination scores.
correlation <- cor(study_hours, exam_scores)

# Display the correlation coefficient.
print(correlation)
```

## Explain

In one or two sentences:

> What does the correlation value tell you about these two variables?

Then answer:

> Does the correlation prove that increasing study hours causes examination scores to increase?

Explain why or why not.

---

# 14. A Complete Mini EDA Workflow

A basic EDA workflow can be organised as:

```text
Load data
   ↓
Inspect structure
   ↓
Check data types
   ↓
Check missing values
   ↓
Summarise variables
   ↓
Visualise distributions
   ↓
Explore relationships
   ↓
Identify possible data problems
   ↓
Interpret findings
```

Therefore:

> **Machine learning does not begin with model training. It begins with understanding the problem and the data.**

## Worked EDA

```{r}
# Create a small data frame for the complete EDA demonstration.
eda_data <- data.frame(
  study_hours = c(2, 3, 4, 5, 6, 7, 8, 9, 10, 11),
  attendance = c(60, 65, 70, 72, 75, 80, 82, 85, 90, 92),
  exam_score = c(50, 55, 60, 65, 68, 72, 78, 82, 88, 91)
)

# Display the first rows to inspect the dataset.
print(head(eda_data))

# Display the dimensions of the dataset.
print(dim(eda_data))

# Display the structure and data types.
str(eda_data)

# Display descriptive statistics.
print(summary(eda_data))

# Count missing values in each column.
print(colSums(is.na(eda_data)))
```

```{r}
# Create a histogram showing the distribution of examination scores.
hist(eda_data$exam_score,
     main = "Distribution of Examination Scores",
     xlab = "Examination Score",
     ylab = "Number of Students")
```

```{r}
# Create a scatter plot showing study hours against examination score.
plot(eda_data$study_hours,
     eda_data$exam_score,
     main = "Study Hours vs Examination Score",
     xlab = "Study Hours",
     ylab = "Examination Score")
```

## EDA Interpretation Task

Based on the output above, write **at least three observations**.

Your observations should be about the data, not about the code.

For example:

- one observation about distribution;
- one observation about a relationship;
- one observation about data quality.

Avoid writing:

> “The code ran successfully.”

That is a programming observation, not a data-science observation.

---

# 15. LAB TASK — Student Performance EDA

## Scenario

You are given a small dataset containing information about students.

Your task is to conduct a basic exploratory analysis before any machine-learning model is considered.

### Objectives

You should be able to:

1. inspect the dataset;
2. identify observations and variables;
3. identify possible features and a target;
4. check the data structure;
5. check missing values;
6. calculate descriptive statistics;
7. create at least two visualisations;
8. examine relationships between numerical variables;
9. communicate at least five findings.

### Important

Do not begin by asking:

> **“Which machine-learning algorithm should I use?”**

Begin by asking:

> **“What does this dataset contain, and what can I learn from it?”**

## Lab Dataset

```{r}
# Create a student-performance dataset for the laboratory exercise.
lab_data <- data.frame(
  study_hours = c(2, 4, 5, 6, 7, 8, 3, 9, 10, 5, 6, 8),
  attendance = c(60, 72, 75, 80, 85, 90, 65, 92, 95, 78, 82, 88),
  assignment_score = c(55, 65, 70, 74, 80, 85, 60, 88, 92, 72, 76, 84),
  exam_score = c(50, 62, 68, 73, 79, 84, 57, 87, 91, 70, 75, 82)
)

# Display the first five observations.
print(head(lab_data))

# Display the number of observations and variables.
print(dim(lab_data))

# Display the data structure.
str(lab_data)

# Display descriptive statistics.
print(summary(lab_data))

# Display the number of missing values in each variable.
print(colSums(is.na(lab_data)))
```

## Lab Task A — Inspect

Answer:

1. What does one row represent?
2. How many observations are there?
3. How many variables are there?
4. Which variable could be the target if the goal is to predict examination performance?
5. Which variables could be features?

## Lab Task B — Visualise

Create your own:

1. scatter plot of `study_hours` against `exam_score`;
2. scatter plot of `attendance` against `exam_score`;
3. boxplot of `exam_score`.

For each plot, write **one sentence explaining what you observe**.

### Starting point: Study Hours vs Exam Score

```{r}
# Create a scatter plot of study hours against examination score.
plot(lab_data$study_hours,
     lab_data$exam_score,
     main = "Study Hours vs Exam Score",
     xlab = "Study Hours",
     ylab = "Exam Score")
```

### Starting point: Attendance vs Exam Score

```{r}
# Create a scatter plot of attendance against examination score.
plot(lab_data$attendance,
     lab_data$exam_score,
     main = "Attendance vs Exam Score",
     xlab = "Attendance",
     ylab = "Exam Score")
```

### Starting point: Exam Score Boxplot

```{r}
# Create a boxplot of examination scores.
boxplot(lab_data$exam_score,
        main = "Exam Score Distribution",
        ylab = "Exam Score")
```

### Correlation Matrix

```{r}
# Calculate the correlation matrix for the numerical variables.
correlation_matrix <- cor(lab_data)

# Display the correlation matrix.
print(correlation_matrix)
```

## Lab Task C — Interpret

Using the correlation matrix and plots:

- Which pair of variables appears most strongly related?
- Which relationship appears weakest?
- Do any observations look unusual?
- Are there missing values?
- What additional information would you want before building a model?

### Lab conclusion

Write a short paragraph answering:

> **What did the EDA tell you that you would not have known from simply looking at the raw table?**

---

# 16. DO IT YOURSELF — Mini EDA Challenge

Now work more independently.

Choose **one** context:

- education;
- agriculture;
- healthcare;
- finance;
- transportation;
- language;
- music/cultural heritage;
- environment.

You may use a small dataset supplied by your lecturer or create a simple dataset yourself.

## Your task

### Step 1 — Define the problem

What question are you trying to answer?

### Step 2 — Identify an observation

What does one row represent?

### Step 3 — Identify variables

Which columns represent information about each observation?

### Step 4 — Identify possible features

Which variables could potentially be used as inputs?

### Step 5 — Identify the target

If this were supervised learning, what would you predict?

### Step 6 — Inspect the data

Use R to examine:

- dimensions;
- column names;
- data types;
- missing values;
- descriptive statistics.

### Step 7 — Visualise

Create at least:

- one distribution plot;
- one relationship plot.

### Step 8 — Interpret

Write at least **five data-driven observations**.

### Step 9 — Identify one concern

Identify one possible:

- missing-data issue;
- unusual value;
- sampling issue;
- representation issue;
- leakage concern.

### Step 10 — Reflect

Explain why you would **not** immediately choose a machine-learning algorithm.

---

# 17. WEEK 0 REFLECTION

Take a few minutes to answer these questions honestly.

## Conceptual Understanding

1. What is the difference between an R vector and an R matrix?
2. What does the `dim()` function tell us?
3. What is a data frame?
4. Why do we inspect a dataset before modelling?
5. What is exploratory data analysis?
6. Why are visualisations useful?
7. What is correlation?
8. Why does correlation not automatically imply causation?

## Practical Confidence

Rate yourself from **1–5**:

| Skill | Rating |
|---|---:|
| R objects and data types | |
| Vectors | |
| Matrices | |
| Lists | |
| Data frames | |
| Selecting and filtering data | |
| Creating new variables | |
| Descriptive statistics | |
| Handling missing values | |
| Creating plots | |
| Interpreting plots | |
| Basic EDA | |

## Final Reflection

Complete these sentences:

> **One thing I can now do confidently is…**

> **One thing I still need to practise is…**

> **One thing I learned about data that surprised me is…**

> **Before building a machine-learning model, I now know that I should first…**

---

# 18. WEEK 0 CHECKLIST

Before moving to Week 1, make sure you can:

- [ ] create and use R objects;
- [ ] work with numeric, character and logical values;
- [ ] create and manipulate vectors;
- [ ] create and inspect matrices;
- [ ] understand rows, columns and dimensions;
- [ ] work with lists;
- [ ] create a data frame;
- [ ] inspect a dataset;
- [ ] select columns and rows;
- [ ] filter observations;
- [ ] create a simple feature;
- [ ] detect missing values;
- [ ] understand that missing data require investigation;
- [ ] calculate basic statistics;
- [ ] create a histogram;
- [ ] create a scatter plot;
- [ ] create a boxplot;
- [ ] calculate and interpret correlation;
- [ ] perform a basic EDA;
- [ ] communicate findings from data.

---

# Bridge to Week 1

You have now learned how to **work with data in R**.

The next question is:

> **How can a computer learn useful patterns from that data?**

That takes us to:

## Week 1 — Introduction to Machine Learning

We will move from:

**Data → EDA**

to:

**Data → Learning → Model → Prediction**

And we will begin with the most important principle:

> **Problem first. Data second. Algorithm later.**
