---
title: "DTS 301 — Machine Learning I"
subtitle: "Week 1: Introduction to Machine Learning"
author: "Dr. Sakinat Folorunso — Associate Professor of AI Systems"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Course position

**Week 0 → Python for Data Science, NumPy, Pandas, Visualization & EDA**  
**Week 1 → Introduction to Machine Learning**  
**Week 2 → Data Representation, Linear Transformations & Matrix–Vector Operations**

> **Learning cycle:** Read → Predict → Run → Observe → Explain → Reflect

## RStudio pathway

This Week 1 practical is designed for **RStudio** and **R Markdown**. Work through the explanations, run the code chunks, inspect the output, and complete the activities in your own words.

# 1. Warm-up: What does it mean for a computer to "learn"?

A lecturer wants to predict whether a student is likely to pass a course.

Available information includes:

- previous test scores;
- attendance;
- assignment performance;
- study hours.

A rule-based approach might be:

```text
IF attendance > 75%
AND test score > 50%
THEN predict PASS
```

Another approach is to provide a computer with many previous student records and their outcomes and allow an algorithm to learn a relationship between the information and the outcome.

### Think before you continue

1. What is different between these two approaches?
2. Which information would the computer need in order to learn from previous examples?
3. What could go wrong if the historical examples are poor or unrepresentative?

### Your response

**Q1. Difference between the two approaches:**


**Q2. Information required:**


**Q3. What could go wrong?**

# 2. Conventional programming versus machine learning

In conventional programming:

\[
\text{Rules} + \text{Data} \rightarrow \text{Output}
\]

In machine learning:

\[
\text{Data} + \text{Examples/Targets} \rightarrow \text{Learned Model}
\]

The learned model can then be used on new data:

\[
\text{New Data} + \text{Learned Model} \rightarrow \text{Prediction}
\]

Machine learning is concerned with methods that learn useful patterns or relationships from data and use what was learned to make predictions or discover structure.

# 3. Machine-learning vocabulary

| Term | Meaning |
|---|---|
| **Observation / instance** | One individual example or record |
| **Variable** | A measurable or recorded property |
| **Feature** | An input variable used by a model |
| **Target / label** | The outcome a supervised model is asked to predict |
| **Dataset** | Collection of observations |
| **Model** | A mathematical/computational representation learned from data |
| **Training** | Process of learning model parameters from data |
| **Prediction** | Model output for an observation |
| **Generalization** | Ability to perform well on unseen data |

### Example

| Study hours | Attendance | Test score | Final result |
|---:|---:|---:|---|
| 5 | 80 | 65 | Pass |
| 2 | 55 | 42 | Fail |
| 7 | 90 | 78 | Pass |

Each row is an **observation**. The first three columns are possible **features**. `Final result` is the **target**.

# 4. From a real problem to an ML problem

Machine learning should begin with a **problem definition**, not with an algorithm.

Consider:

> Can we predict whether a loan applicant will repay a loan?

Ask:

1. What is one observation?
2. What information is available when the prediction is made?
3. What is the target?
4. Is the target categorical or numerical?
5. What would a useful prediction look like?
6. What are the consequences of a wrong prediction?

### Student activity

For the loan example, complete:

- Observation =
- Possible features =
- Target =
- Type of task =
- One ethical/data-quality concern =

# 5. The Iris dataset: our first ML dataset

We use the classic Iris dataset. Each observation describes an iris flower using four measurements:

- sepal length;
- sepal width;
- petal length;
- petal width.

The target identifies the flower species.

```{r}
# Load the built-in Iris dataset.
data(iris)

# Display the first six observations.
head(iris)

# Display the dimensions of the dataset.
dim(iris)
```

## Observe

The dataset contains **150 observations**, **4 input features**, and **1 target column**.

In R, the four measurement columns form the input feature matrix \(X\), while `Species` is the target.

```{r}
# Select the four numerical feature columns.
X <- iris[, 1:4]

# Select the target column.
y <- iris$Species

# Display the dimensions of X.
dim(X)

# Display the dimensions of y as a vector.
length(y)

# Display the feature names.
names(X)

# Display the target classes.
levels(y)
```

# 6. Dataset representation

If there are \(n\) observations and \(p\) features:

\[
X \in \mathbb{R}^{n \times p}
\]

For Iris:

\[
X \in \mathbb{R}^{150 \times 4}
\]

Each observation can be represented as a feature vector:

\[
x_i =
\begin{bmatrix}
x_{i1}\\
x_{i2}\\
\vdots\\
x_{ip}
\end{bmatrix}
\]

For example, the first iris observation is:

```{r}
# Extract the first observation as a numerical feature vector.
first_observation <- as.numeric(X[1, ])

# Display the feature vector.
first_observation

# Count the number of features.
length(first_observation)
```

The feature vector is a numerical representation of one observation.

This is the bridge between the data-analysis skills from Week 0 and the mathematical representation of machine learning that we develop further in Week 2.

# 7. Classification

Classification is a supervised-learning task in which the target represents a category or class.

Iris has three classes:

\[
\{\text{setosa},\text{versicolor},\text{virginica}\}
\]

Unlike numerical codes, these names make it clear that the categories are not quantities.

```{r}
# Count the observations in each Iris species.
table(y)

# Display the class names.
levels(y)
```

### Think

Would you call Iris a classification problem or a regression problem?

**Why?**

### Answer

Iris is a **classification** problem because the target identifies one of a finite set of categories/classes.

# 8. Regression

Regression is a supervised-learning task in which the target is typically numerical.

Examples include predicting:

- house price;
- rainfall amount;
- temperature;
- electricity demand;
- student examination score.

A simplified formulation is:

\[
y \in \mathbb{R}
\]

For example:

\[
\text{score}=72.5
\]

is a numerical target rather than a category.

| Problem | Target | ML task |
|---|---|---|
| Predict flower species | Category | Classification |
| Predict house price | Number | Regression |
| Predict rainfall amount | Number | Regression |
| Predict disease class | Category | Classification |

# 9. Unsupervised learning

In supervised learning, we have input \(X\) and target \(y\).

In unsupervised learning, we generally have input data \(X\) without a supplied target.

The goal may be to discover:

- groups;
- patterns;
- structure;
- lower-dimensional representations;
- unusual observations.

For example:

> Can we group customers according to similarities in their behaviour?

We will study clustering and dimensionality reduction later in the course.

# 10. Supervised learning mathematically

Suppose our dataset is:

\[
D=\{(x_i,y_i)\}_{i=1}^{n}
\]

where \(x_i\) is the feature vector for observation \(i\), and \(y_i\) is the corresponding target.

A supervised-learning algorithm attempts to learn:

\[
\hat{y}=f_{\theta}(x)
\]

where:

- \(x\) = input features;
- \(y\) = actual target;
- \(\hat{y}\) = predicted target;
- \(f_{\theta}\) = learned model;
- \(\theta\) = model parameters.

In plain language, we want the computer to learn a useful relationship between the inputs and target so it can make predictions for new observations.

# 11. Prediction and loss

Suppose the true examination score is:

\[
y=70
\]

Two models make these predictions:

- Model A: \(\hat y=64\)
- Model B: \(\hat y=69\)

One simple loss function is squared error:

\[
L(y,\hat y)=(y-\hat y)^2
\]

```{r}
# Store the actual examination score.
actual <- 70

# Store Model A's prediction.
prediction_A <- 64

# Store Model B's prediction.
prediction_B <- 69

# Calculate Model A's squared error.
loss_A <- (actual - prediction_A)^2

# Calculate Model B's squared error.
loss_B <- (actual - prediction_B)^2

# Display both losses.
loss_A
loss_B
```

### Think

Which prediction is closer to the true value?

Which model has the smaller squared error?

A loss function quantifies how different a prediction is from the actual target.

# 12. Empirical risk

For \(n\) observations, average loss can be written as:

\[
\hat R(\theta)
=
\frac{1}{n}
\sum_{i=1}^{n}
L(f_{\theta}(x_i),y_i)
\]

This is one way to express **empirical risk**.

The broad learning idea is to find model parameters that perform well according to an appropriate objective on training data while still aiming for good performance on unseen data.

# 13. Training versus generalization

A model can perform extremely well on examples it has already seen. That does not automatically mean it will perform well on new observations.

### Training performance

How well does the model perform on data used during learning?

### Generalization

How well does the model perform on data it did not see during training?

The goal is not simply to memorize the training data. The goal is to learn patterns that remain useful for new observations.

### Analogy

Imagine a student memorizes the exact answers to 20 practice questions. The student scores 100% when those exact questions appear again but performs poorly on new questions.

A model can face a similar problem.

# 14. Training, validation and test data

A common workflow separates data into different roles:

```text
Dataset
   |
   +---- Training set
   |       ↓
   |    Learn model
   |
   +---- Validation set
   |       ↓
   |    Tune/select
   |
   +---- Test set
           ↓
      Final evaluation
```

The exact workflow can vary depending on the dataset, algorithm and experimental design.

The key principle is:

> **Evaluation data should provide an honest estimate of performance on unseen data.**

# 15. EDA before modelling

Before training a model, inspect:

- dimensions;
- data types;
- missing values;
- distributions;
- relationships;
- unusual observations.

Let's perform a quick EDA on Iris.

```{r}
# Display the dimensions of the Iris dataset.
dim(iris)

# Count missing values in each column.
colSums(is.na(iris))

# Display summary statistics for the dataset.
summary(iris)
```

## Visualisation

```{r}
# Create a scatter plot of sepal length against petal length.
plot(
  iris$Sepal.Length,
  iris$Petal.Length,
  xlab = "Sepal length (cm)",
  ylab = "Petal length (cm)",
  main = "Iris: Sepal Length vs Petal Length"
)
```

### Interpret the plot

Write at least **two observations**.

1.
2.
3.

# 16. Data leakage

Data leakage occurs when information that would not legitimately be available at prediction time enters the modelling process.

Suppose we want to predict whether a student will graduate **before graduation**.

Potential features include:

- first-year GPA;
- attendance;
- assignment performance;
- number of completed courses.

But including **final graduation status** as a feature would be inappropriate because it is the outcome itself.

### Key question

> **Would this information genuinely be available at the moment the prediction is supposed to be made?**

If not, we may have leakage.

# 17. Responsible machine learning

Machine learning systems are built from data. Technical performance is therefore not the only issue.

### Data

- Who collected it?
- How was it collected?
- Who is represented?
- Who is missing?

### Target

- Is the target clearly defined?
- Is it measured reliably?
- Does it represent what we actually care about?

### Features

- Are the features available at prediction time?
- Could they contain leakage?
- Are they appropriate for the task?

### Consequences

- What happens when the model is wrong?
- Who is affected?
- How will predictions be used?

# 18. Student-centred activity: ML Problem Detective

For each scenario:

1. Identify one observation.
2. Identify possible features \(X\).
3. Identify the target \(y\), if there is one.
4. Decide whether it is classification, regression or unsupervised learning.
5. Identify one data-quality, leakage or deployment concern.

### Scenario A — Fraud detection

A bank wants to identify whether a transaction is potentially fraudulent.

### Scenario B — House prices

A real-estate company wants to predict the selling price of a house.

### Scenario C — Customer segmentation

A retailer wants to group customers according to purchasing behaviour, but has no predefined customer-group labels.

### Scenario D — Yoruba emotion classification

A research team wants to classify Yoruba-language text into predefined emotional categories.

### Scenario E — Rainfall prediction

A meteorological team wants to predict rainfall amount using weather measurements.

## Group discussion record

### Scenario A

Observation:

Features:

Target:

Task type:

Concern:

### Scenario B

Observation:

Features:

Target:

Task type:

Concern:

### Scenario C

Observation:

Features:

Target:

Task type:

Concern:

### Scenario D

Observation:

Features:

Target:

Task type:

Concern:

### Scenario E

Observation:

Features:

Target:

Task type:

Concern:

# 19. Mathematical formulation task

Choose one scenario.

Write the problem using:

\[
\hat y=f_{\theta}(x)
\]

Then explain:

- What is \(x\)?
- What is \(y\)?
- What does \(\hat y\) represent?
- What would the learned model \(f_{\theta}\) do?
- What would count as a useful prediction?

**Scenario:**


**\(x\):**


**\(y\):**


**\(\hat y\):**


**\(f_{\theta}\):**


**Explanation:**

# 20. Iris mini-practical

Work independently.

### Task 1

State:

- number of observations;
- number of features;
- number of target classes.

### Task 2

Print the dimensions of \(X\) and the length of \(y\).

### Task 3

Display one feature vector.

### Task 4

Create a visualization involving two features.

### Task 5

Write **three EDA observations**.

### Task 6

Explain why Iris is a classification problem.

### Task 7

Explain the difference between a feature and a target.

```{r}
# Your work starts here.

# Task 1


# Task 2


# Task 3


# Task 4


# Task 5


# Task 6


# Task 7
```

# 21. Class Task / Do It Yourself

## ML Problem Formulation Challenge

Choose a real problem from:

- education;
- agriculture;
- healthcare;
- finance;
- transportation;
- language;
- music/culture;
- environment.

**Do not start by choosing an algorithm. Start with the problem.**

Complete:

1. What is one observation?
2. What are the possible features?
3. What is the target, if there is one?
4. Is this classification, regression or unsupervised learning?
5. Write the mathematical abstraction \(\hat y=f_{\theta}(x)\).
6. What information would be unavailable at prediction time?
7. Identify one possible source of data leakage.
8. Identify one possible data-quality concern.
9. What would a wrong prediction mean in the real world?

### Deliverable

**Problem → Data → Features → Target → ML task → Mathematical formulation → Risk/concern**

# 22. Reflection

Answer these questions honestly.

1. In your own words, what is machine learning?
2. What is the difference between conventional programming and machine learning?
3. What is an observation?
4. What is a feature?
5. What is a target?
6. What is the difference between classification and regression?
7. When would unsupervised learning be appropriate?
8. What does \(X \in \mathbb{R}^{n\times p}\) mean?
9. What does \(\hat y=f_{\theta}(x)\) mean?
10. Why is generalization more important than simply memorizing training examples?
11. Why is EDA important before modelling?
12. What is data leakage?
13. What question will you ask yourself before including a feature in a model?

### Exit ticket

> **The most important idea I learned this week is __________ because __________.**

# 23. Week 1 checklist

Before moving to Week 2, check yourself.

- [ ] I can explain machine learning in my own words.
- [ ] I can distinguish conventional programming from machine learning.
- [ ] I understand observation, variable, feature and target.
- [ ] I can distinguish classification and regression.
- [ ] I understand the basic idea of unsupervised learning.
- [ ] I can represent a dataset as \(X\) and \(y\).
- [ ] I understand \(X \in \mathbb{R}^{n\times p}\).
- [ ] I understand the meaning of \(\hat y=f_{\theta}(x)\).
- [ ] I understand the purpose of a loss function.
- [ ] I understand training versus unseen data.
- [ ] I understand the basic idea of generalization.
- [ ] I know why EDA comes before modelling.
- [ ] I can identify a simple data-leakage problem.
- [ ] I can formulate a real-world problem as an ML problem.

# 24. Bridge to Week 2

You can now describe a machine-learning problem in terms of:

\[
X,\quad y,\quad x_i,\quad \hat y,\quad f_{\theta}
\]

But we need to go deeper.

How exactly do we represent information mathematically?

How do vectors and matrices transform data?

What does multiplication between matrices and vectors mean?

How can transformations change the representation of our data?

These questions lead directly to:

# Week 2 — Data Representation, Linear Transformations & Matrix–Vector Operations

Your R skills from Week 0 will become the computational tools for exploring these mathematical ideas.

---

## Instructor contact

**Dr. Sakinat Folorunso**  
Associate Professor of AI Systems  
Department of Computer Science, Olabisi Onabanjo University  
**Email:** sakinat.folorunso@oouagoiwoye.edu.ng

**DTS 301 — Machine Learning I**
