---
title: "DTS 301 — Machine Learning I: Week 2"
subtitle: "Data Representation, Features, Feature Vectors & Matrix–Vector Operations"
author: "Dr. Sakinat Folorunso"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Week 2

**Instructor:** Dr. Sakinat Folorunso  
**Department:** Computer Science, Olabisi Onabanjo University (OOU)

## Learning Hours (LH)

Representation of data; features; feature vectors; linear transformations; matrix and vector operations in data representation.

## Practical Hours (PH)

Create feature matrices and vectors; perform basic matrix-vector operations; inspect and transform a small tabular dataset.

---

## How to use this workbook

This is a **student-centered working workbook** for RStudio.

Use the learning cycle:

**Read → Predict → Run → Inspect → Explain → Reflect**

For every worked example:

1. Read the problem.
2. Predict what should happen.
3. Run the R code.
4. Inspect the output.
5. Explain what the output means.
6. Reflect on how the idea connects to machine learning.

> **Core idea:** Before an algorithm can learn, the data must be represented in a form that the algorithm can use.

# 1. Week 1 → Week 2

In Week 1, we moved from **working with data** to **learning from data**.

We introduced observations, features, targets, classification, regression, supervised learning, training, generalization, and the idea of a dataset as inputs and outputs.

This week we ask:

**How exactly do we represent observations and features mathematically?**

The progression is:

**Problem → Data → Features → Representation → Mathematical Operations → Prediction**

# 2. Learning outcomes

By the end of this workbook, you should be able to:

1. Identify features in a dataset.
2. Construct a feature vector for one observation.
3. Construct a feature matrix for several observations.
4. Explain the meaning of rows and columns in a feature matrix.
5. Perform vector addition, subtraction, scalar multiplication and dot products.
6. Perform matrix–vector multiplication.
7. Check whether matrix dimensions are compatible for multiplication.
8. Describe a linear transformation.
9. Apply a simple transformation to a small dataset.
10. Explain why representation matters before model training.

# 3. From a real-world observation to data

Consider:

| Student | Study Hours | Attendance | Previous Score |
|---|---:|---:|---:|
| A | 6 | 85 | 68 |
| B | 8 | 92 | 75 |
| C | 4 | 70 | 55 |

For Student A, the numerical information can be represented as:

**x = (6, 85, 68)**

This is a **feature vector**.

## R example: one feature vector

```{r}
# Create one feature vector for Student A.
x <- c(6, 85, 68)

# Display the feature vector.
x

# Display the number of values in the feature vector.
length(x)
```

### Observe

The vector contains three values, so Student A is represented using **three features**.

# 4. Feature vectors

For an observation i with p features:

**xᵢ = (xᵢ₁, xᵢ₂, …, xᵢₚ)**

For Student A:

**x = (6, 85, 68)**

The important idea is that an observation is converted into a numerical representation that mathematical operations can use.

## Iris example

The Iris dataset has four numerical features:

- sepal length
- sepal width
- petal length
- petal width

One flower can therefore be represented by a four-dimensional feature vector.

```{r}
# Create a hypothetical Iris feature vector with four measurements.
iris_x <- c(5.1, 3.5, 1.4, 0.2)

# Display the feature vector.
iris_x

# Count the number of features.
length(iris_x)
```

# 5. Feature matrices

When we have several observations, their feature vectors can be placed together to form a **feature matrix**.

For n observations and p features:

**X ∈ ℝⁿˣᵖ**

Rows represent observations.  
Columns represent features.

For the student dataset, there are three observations and three features:

**X ∈ ℝ³ˣ³**

## R example: create a feature matrix

```{r}
# Create the feature matrix for the three students.
X <- matrix(
  c(
    6, 85, 68,
    8, 92, 75,
    4, 70, 55
  ),
  nrow = 3,
  byrow = TRUE
)

# Display the feature matrix.
X

# Display the dimensions of the matrix.
dim(X)

# Display the number of observations.
nrow(X)

# Display the number of features.
ncol(X)
```

## Check your understanding

If a dataset contains **500 observations and 10 features**, its feature matrix has shape:

**500 × 10**

In mathematical notation:

**X ∈ ℝ⁵⁰⁰ˣ¹⁰**

# 6. Inspecting rows and columns

Rows correspond to observations.

Columns correspond to features.

```{r}
# Select the first row, representing Student A.
student_A <- X[1, ]

# Display Student A's feature vector.
student_A

# Select the first column, representing study hours.
study_hours <- X[, 1]

# Display study hours for all students.
study_hours

# Select the second column, representing attendance.
attendance <- X[, 2]

# Display attendance for all students.
attendance
```

## Add meaningful column names

```{r}
# Assign descriptive names to the feature columns.
colnames(X) <- c("Study_Hours", "Attendance", "Previous_Score")

# Display the labelled feature matrix.
X
```

# 7. Vector operations

Let:

**a = (2, 4, 6)**

**b = (1, 3, 5)**

We can perform:

- addition
- subtraction
- scalar multiplication
- dot product

```{r}
# Create vector a.
a <- c(2, 4, 6)

# Create vector b.
b <- c(1, 3, 5)

# Add the vectors element by element.
a + b

# Subtract vector b from vector a.
a - b

# Multiply vector a by the scalar value 2.
2 * a

# Calculate the dot product using the sum of element-wise products.
sum(a * b)
```

## Dot product

For:

**a = (2, 4, 6)**

**b = (1, 3, 5)**

the dot product is:

**a · b = (2×1) + (4×3) + (6×5) = 44**

### Predict

What should this give?

**(3, 2, 1) · (1, 2, 3)**

```{r}
# Create the first prediction vector.
u <- c(3, 2, 1)

# Create the second prediction vector.
v <- c(1, 2, 3)

# Calculate the dot product.
sum(u * v)
```

# 8. Matrix–vector multiplication

Consider:

**A = [[1, 2], [3, 4]]**

and:

**x = (5, 6)**

The multiplication gives:

**Ax = (17, 39)**

In R, matrix multiplication uses `%*%`.

```{r}
# Create the 2 by 2 matrix A.
A <- matrix(
  c(
    1, 2,
    3, 4
  ),
  nrow = 2,
  byrow = TRUE
)

# Create the two-element vector x.
x <- c(5, 6)

# Multiply matrix A by vector x.
Ax <- A %*% x

# Display the result.
Ax

# Display the dimensions of the result.
dim(Ax)
```

# 9. Matrix dimension compatibility

Matrix multiplication is possible when the **inner dimensions match**.

For example:

`(3 × 2)(2 × 1)` is valid.

The result has shape:

`3 × 1`

But:

`(3 × 2)(3 × 1)` is not valid because the inner dimensions do not match.

For:

`(m × n)(n × p)`

the result has shape:

`m × p`

## R example

```{r}
# Create a 3 by 2 matrix.
M <- matrix(
  c(
    1, 2,
    3, 4,
    5, 6
  ),
  nrow = 3,
  byrow = TRUE
)

# Create a compatible two-element vector.
v2 <- c(10, 20)

# Multiply the matrix by the vector.
result <- M %*% v2

# Display the result.
result

# Display the dimensions of the result.
dim(result)
```

# 10. Linear transformations

A linear transformation changes a vector using a matrix.

Consider:

**A = [[2, 0], [0, 2]]**

and:

**x = (2, 3)**

Then:

**Ax = (4, 6)**

The transformation doubles both coordinates.

```{r}
# Create a simple scaling matrix.
scale <- matrix(
  c(
    2, 0,
    0, 2
  ),
  nrow = 2,
  byrow = TRUE
)

# Create a two-dimensional input vector.
point <- c(2, 3)

# Apply the linear transformation.
transformed_point <- scale %*% point

# Display the original vector.
point

# Display the transformed vector.
transformed_point
```

# 11. A simple feature transformation

Features can have different numerical scales.

For example:

- study hours may range from 0 to 20
- attendance may range from 0 to 100
- scores may range from 0 to 100

One simple transformation is:

**x′ = x / 100**

This is only a simple demonstration. Later topics will introduce more systematic feature-scaling methods.

```{r}
# Create a copy of the feature matrix so the original remains unchanged.
X_scaled <- X

# Scale the attendance column from 0–100 to 0–1.
X_scaled[, "Attendance"] <- X_scaled[, "Attendance"] / 100

# Scale the previous-score column from 0–100 to 0–1.
X_scaled[, "Previous_Score"] <- X_scaled[, "Previous_Score"] / 100

# Display the transformed matrix.
X_scaled
```

## Inspect and explain

Ask yourself:

1. Which columns changed?
2. Which column did not change?
3. Why might transforming features be useful?
4. Does transforming a feature automatically make the data better?

Write your explanation in your own words.

# 12. Class Task — Representation Detective

Work in pairs.

Choose one scenario:

### Healthcare
Possible features:
- age
- blood pressure
- glucose level
- previous diagnosis

### Education
Possible features:
- study hours
- attendance
- assignment score
- previous examination score

### Agriculture
Possible features:
- rainfall
- temperature
- soil moisture
- fertilizer quantity

### Music
Possible features:
- tempo
- spectral features
- duration
- other extracted audio features

For your chosen scenario:

1. Identify one observation.
2. Identify at least three features.
3. Write a possible feature vector.
4. Explain what one row of the feature matrix represents.
5. Explain what one column represents.

# 13. Do It Yourself

Use this dataset:

| Student | Study Hours | Attendance | Assignment | Previous Score |
|---|---:|---:|---:|---:|
| A | 5 | 80 | 72 | 65 |
| B | 8 | 90 | 85 | 78 |
| C | 3 | 65 | 55 | 50 |
| D | 7 | 88 | 80 | 74 |

## Task 1 — Create the feature matrix

```{r}
# Create the feature matrix for the four students.
X_task <- matrix(
  c(
    5, 80, 72, 65,
    8, 90, 85, 78,
    3, 65, 55, 50,
    7, 88, 80, 74
  ),
  nrow = 4,
  byrow = TRUE
)

# Give the feature columns meaningful names.
colnames(X_task) <- c(
  "Study_Hours",
  "Attendance",
  "Assignment",
  "Previous_Score"
)

# Display the feature matrix.
X_task

# Display its dimensions.
dim(X_task)
```

## Task 2 — Extract Student B

```{r}
# Extract the second row, representing Student B.
student_B <- X_task[2, ]

# Display Student B's feature vector.
student_B
```

## Task 3 — Extract attendance

```{r}
# Extract the attendance column.
attendance_task <- X_task[, "Attendance"]

# Display attendance values.
attendance_task
```

## Task 4 — Dot product of the first two students

```{r}
# Extract Student A's feature vector.
student_A_task <- X_task[1, ]

# Extract Student B's feature vector.
student_B_task <- X_task[2, ]

# Calculate the dot product.
sum(student_A_task * student_B_task)
```

## Task 5 — Transform attendance and assignment

```{r}
# Create a copy of the task matrix.
X_task_scaled <- X_task

# Scale attendance to a 0–1 range.
X_task_scaled[, "Attendance"] <- X_task_scaled[, "Attendance"] / 100

# Scale assignment to a 0–1 range.
X_task_scaled[, "Assignment"] <- X_task_scaled[, "Assignment"] / 100

# Display the transformed matrix.
X_task_scaled
```

### Explain

Why does the transformed matrix still represent the same students and the same features even though some numerical values changed?

# 14. Reflection

Complete these statements:

1. A feature is ____________________________________________.
2. A feature vector represents ______________________________.
3. In a feature matrix, rows represent _______________________.
4. In a feature matrix, columns represent ____________________.
5. Matrix–vector multiplication is valid when ________________.
6. A linear transformation can ______________________________.
7. Data representation matters because _______________________.

## One-minute exit question

If a machine-learning dataset has **1,000 observations and 12 features**, what is the shape of its feature matrix?

**Your answer:** __________________

# 15. Week 2 checklist

Before moving on, check that you can:

- [ ] identify features in a dataset
- [ ] construct a feature vector
- [ ] construct a feature matrix
- [ ] explain rows and columns
- [ ] inspect matrix dimensions in R
- [ ] perform vector addition and subtraction
- [ ] perform scalar multiplication
- [ ] calculate a dot product
- [ ] perform matrix–vector multiplication
- [ ] check matrix dimensions
- [ ] explain a simple linear transformation
- [ ] apply a simple feature transformation
- [ ] explain why representation matters

# Bridge to Week 3

You now know how data becomes a numerical representation.

Next we ask:

**What kinds of prediction problems can we formulate from these representations?**

## Week 3

**Classification, Regression, Problem Formulation & Probability**

The progression continues:

**Problem → Data → Features → Representation → Prediction Problem → Model → Evaluation**
